# Chapter 1. The fate of South America's endemic mammalian fauna in response to the most dramatic Cenozoic climate disruption {.unnumbered header-title="Chapter 1"}

## Foreword

```{r, echo=FALSE}
library(knitr)
```

Understanding what drove the emergence and demise of lineages during biodiversity crises may provide key insights at the time of the sixth mass extinction. Here, using a vetted fossil database, we explored how South American mammals (SAM) responded to the most dramatic Cenozoic climate cooling episode, the Eocene--Oligocene transition (ca. 34 Ma). We found that SAM experienced a gradual and long-standing diversity decline strongly related to late Eocene progressive climate cooling, followed by a major taxonomic turnover in the Oligocene, driven by diversity-dependent factors, Andean uplift and perhaps grassland expansion. SAM were therefore more subject to long-term diversity changes rather than abrupt ones, contrary to what would be expected in a mass extinction scenario.

The content of this chapter was published in an international peer-reviewed journal [@buffan2025].

## Abstract

Around 34 million years ago (Ma), the Eocene--Oligocene transition (EOT) marked the most dramatic global climatic cooling of the Cenozoic. On a planetary scale, palaeontological evidence suggests that this transition was associated with major faunal turnovers, sometimes even regarded as a mass extinction crisis. In South America, there is no consensus on the response of the endemic mammals to this transition. Here, using a vetted fossil dataset and cutting-edge Bayesian methods, we analyzed the dynamics of South American mammal (SAM) diversification and their possible drivers across latitude (tropical vs. extratropical), taxonomic groups, and trophic guilds throughout the Eocene--Oligocene (ca. 56--23 Ma). Our results did not evidence any mass extinction among SAM at the EOT. Instead, they experienced a gradual and long-term diversity decline from the middle Eocene to the early Oligocene, followed by a sudden waxing-and-waning diversity associated with a large taxonomic -- but not ecological -- turnover. Tropical and extratropical lineages have had very distinct macroevolutionary histories. No effective change in the pace at which tropical lineages diversify was found, thus favoring the tropical stability hypothesis proposed by Wallace. Diversity-dependent effects, temperature and Andean uplift were recovered as probable drivers of SAM diversification across the period. Contrasting evidence casts doubt on the common hypothesis primarily linking Oligocene faunal changes to grassland expansion. Our findings illustrate the uniqueness of the deep-time interplay between endemic SAM and their physical environment in a context of climatic shift, highlighting the need to consider regional idiosyncrasies for understanding the coevolution of life and climate.

## Introduction

Today, South America is home to a unique mammalian fauna, hosting one of the most species-rich regions of the world, Amazonia [@willig2003]. Although the extant mammalian taxonomic diversity in the continent is astonishing, it is a poor reflection of what can be drawn from the fossil record. In fact, throughout most of the Cenozoic (i.e., the last 66 million years), South America was isolated from other landmasses [@scotese_atlas_2021]. In this insular context, terrestrial mammal lineages that inhabited this huge landmass experienced unique radiations [@goin2016; @boivin2019; @croft2020; @beck2023; @tejada2024]. According to @simpson1980 , South American mammal (SAM) communities through the Cenozoic are well summarized by three successive 'faunal strata'. The first one, stretching from the early palaeocene (66 million years ago, Ma) to the late Eocene (ca. 37 Ma), is characterized by the early establishment and radiation of endemic lineages, mostly xenarthrans, metatherians and native ungulates. Then, from the early Oligocene (ca. 33 Ma) until the Late Miocene-Early Pliocene (ca. 5 Ma), key morphological innovations appeared among native lineages, leading to a so-called 'faunal modernization' [@croft2008]. In the meantime, rodents and primates, Afro-Asian lineages that most likely reached South America in the middle-late Eocene [@antoine2012; @bond2015; @marivaux2023], diversified locally [@boivin2019; @beck2023]. In Plio-Pleistocene times, this endemic fauna was massively disrupted. Meanwhile, the closure of the Panama isthmus [@montes2015] had paved the way for major faunal exchanges between North and South America, described as the 'Great American Biotic Interchange' [@marshall1988], and immigrant taxa from North America (*e*.*g*., carnivores, perissodactyls, artiodactyls, or proboscideans) appeared in the South American fossil record. In the last millions of years, many endemic South American lineages declined, or even went totally extinct (e.g., litopterns, notoungulates, sparassodonts, glyptodonts, ground sloths).

Close to the limit between Simpson's first and second faunal strata, the Eocene--Oligocene transition (EOT, *ca*. 33.9 Ma) marked the most dramatic Cenozoic climatic change episode worldwide [@coxall2007]. At that time, simultaneously, global climate massively cooled down [@westerhold2020], ocean level dropped [@miller2020], and the whole Antarctic continent turned into a giant ice sheet [@liu2009]. The prevailing mechanism proposed to explain these major environmental changes is the onset of the Antarctic circumpolar current [@toumoulin2020]. Concomitantly, a major biotic turnover among mammals, known as the '*Grande Coupure*', occurred in Europe [@stehlin1909]. This event, punctuated by the arrival of immigrant taxa from Asia, saw the demise of many endemic European lineages [@legendre1991; @escarguel2008; @weppe2023]. Although European-oriented case studies dominate the literature about the topic, similar disruptions among mammalian faunas occurred at the same time in other regions of the world, such as Asia [@meng1998; @kraatz2010; @sun_synchronous_2014], North America [@figueirido2012; @stucky2014], Africa [@devries2021] or to some extent Madagascar [@godfrey2020]. Recently, @hoyalcuthill2020 even came to rank this geological boundary as the second-most important extinction event throughout the history of life.

In central and southern South America, substantial changes have been documented in metatherians [@abello2020; @goin2010] and native ungulates [@croft2020; @solórzano2024] near the EOT. As an echo to the '*Grande Coupure*', @goin2010 even introduced the term '*Bisagra Patagónica*' (Patagonian Hinge) to characterize the changes among metatherian communities in Patagonia possibly related to the EOT. General morphological trends have been highlighted in 'modernised' notoungulate faunas, with a net increase in body mass and cheek teeth's crown height [@solórzano2021]. The latter changes, by analogy with the evolutionary fate of extant groups, suggested the emergence of the grazing feeding mode, and was postulated to result from the expansion of grasslands, promoted by post-EOT climate cooling and aridification [@solórzano2024; @stebbins1981; @reguero2010]. In northern South America, changes in methaterian and rodent communities around the EOT have also been suggested despite the paucity of known Palaeogene low-latitude fossil localities to date [@antoine2021]. Additionally, geological, palaeontological and phylogenetic evidence suggests that the Eocene--Oligocene was a key interval for the dispersal of SAM toward Caribbean Islands, further playing a central role in the architecture of these unique insular ecosystems [@tejada2024; @philippon2020; @cornée2021; @marivaux_early_2020].

Despite general agreement that a faunal remodelling occurred in SAM communities near the EOT, the evolutionary tempo and mode remain unknown. At macroevolutionary scales, as we would expect in a mass extinction scenario (*SI Appendix*, *Supplementary text*), it is unclear whether this faunal renewal was characterized by extinction of Eocene lineages followed by diversification of the surviving ones [@guinot2023], or whether lineages gradually transitioned until they became 'modernised'. A crucial question underlying these conflicting macroevolutionary hypotheses is whether or not extinction occurred in the tropics, and at what level. Indeed, it has long been postulated that the climatic stability of tropics would make tropical lineages less vulnerable to extinction than extratropical lineages [@wallace1878]. Also, the ecological selectivity and the likely drivers of this apparent faunal transition have remained underexplored. For instance, there is currently no consensus regarding the role of temperature [@kohn2004; @kohn2015] or grassland expansion [@strömberg2013] in shaping the macroevolutionary fate of SAM. Finally, there is a lack of integrative frameworks at the continental scale, with most case studies at the local scale, mostly in southern South America. Nevertheless, the Palaeogene mammal fossil record in South America is spatially and temporally uneven. Spatial gaps cover large areas, such as most of the Amazon basin, and important time intervals have remained unsampled to date (*SI Appendix*, *Figure S1* and *S3*). Accordingly, it has been historically challenging to address questions regarding the history of SAM at the continental scale.

Here, we aim to fill these gaps by examining the available fossil record of SAM. We assembled a genus-level fossil dataset consisting of 3,378 occurrences (distributed in 1,115 species, 536 genera, 96 families, and 18 orders) spanning from the late palaeocene to the Early Miocene. Relying on a fossil-based Bayesian birth-death framework accounting for variable preservation rates through time and across taxa [@silvestro_bayesian_2014], we estimated the tempo and modes of the mammalian diversification dynamics in South America throughout the Eocene--Oligocene interval (*ca*. 56--23 Ma). We then disentangled the underlying taxonomic, functional, and geographic (tropical vs. extratropical) bases of the inferred continent-scale patterns of diversity and diversification among SAM faunas. We finally quantified the most likely biotic and abiotic drivers explaining their diversification dynamics.

## Results and discussion

### No mass extinction: South American mammal diversity gradually declined well before the EOT

We analyzed the fossil database (Table S1) both at the genus and species levels using the process-based Bayesian framework of diversification accounting for varying preservation implemented in PyRate [@silvestro_bayesian_2014]. Our results highlight that the overall mammal diversity gradually declined from the end of the early Eocene -- shortly after the Early Eocene Climate Optimum (EECO, 45) -- to the end of the early Oligocene (decline by respectively 34% and 52% in the number of genera and species between 48 and 32 Ma), and then further increased in the late Oligocene (@fig-DivRTT D and SI Appendix, Figure S8-11). No massive loss of lineages was evidenced at the EOT. This diversity trajectory results from a significant decrease in origination rate during the early Eocene (@fig-DivRTT B and SI Appendix, Figure S8-11). Additionally, we recover a joint significant increase in the pace of origination and extinction (i.e., increased turnover) during the late Eocene, with the extinction rate always being higher than the origination rate (slightly negative net diversification rate; @fig-DivRTT B--C and SI Appendix, S8-11). Similar diversification patterns are recovered when analyzing the dataset at the genus or species level (SI Appendix, Figure S8-11), after removing singletons (SI Appendix, Figure S12-13) or disabling several occurrences of the same lineage that co-occur temporally and geographically (SI Appendix, Figure S14-15). Moreover, general trends in diversity and diversification dynamics are conserved when working with occurrences assigned to South American Land Mammal Ages (SALMAs; SI Appendix, Figure S9 and S11) instead of sub-epochs (see Methods; SI Appendix, Figure S1-2 and Extended Material and Methods), or when considering attempts to derive early Oligocene ages to western Amazonian occurrences originally assigned to an Eocene age (SI Appendix, Figure S32).

Recently, an analysis of the global Phanerozoic fossil record classified the EOT as the second most intense mass extinction episode [@hoyalcuthill2020] (see SI Appendix, Supplementary text for a definition of a mass extinction). Although this ranking was unexpectedly high, drastic extinctions at the EOT are supported by analyses of the European [@stehlin1909] and Asian [@meng1998] mammal fossil record. Applying a methodological framework similar to ours, @weppe2023 recovered a strong extinction signal among endemic Western European artiodactyls at the EOT, with the loss of 62% of genera and 77% of species in only 1 My. Other studies working on mass extinctions illustrated similarly high levels of lineage loss for various study models, for instance in insects at the Permian/Triassic boundary (82% genus loss) [@jouault_multiple_2022] or in elasmobranchs – sharks, skates and rays – at the Cretaceous/palaeogene boundary (44.94% genus and 62.6% species loss) [@guinot2023]. Here, given the relatively small number of lineages lost at the Eocene--Oligocene boundary -- 15 genera (19%), 19 species (22%) --, our results are not consistent with a mass extinction event among SAM at the EOT. Instead, we highlight a long-term decline in SAM diversity, from the middle Eocene onward, contrasting with the expected short temporality of a mass extinction crisis. To our knowledge, our work is the first to provide quantitative support for such a pattern in SAM. These differences between South America and other continents could come from the tropical nature of and the ecological continuum within a significant part of the landmass. Stemming from the cornerstone hypothesis of Wallace linking the striking similarities between tropical biomes from all-over the world to the uniformity of tropical climate across space and (shallow) time [@wallace1878], tropical climate stability would act as a buffer for tropical lineages against extinction. Also, the isolated condition of the landmass has been proposed as a probable cause explaining the singularity of SAM near-EOT macroevolutionary fate, as it greatly limited the possibilities for immigration events. Yet, immigrations were hypothesized to have played a key role in EOT-related extinctions in other parts of the world [@goin2016] [but see @escarguel2008].

```{r, echo=FALSE}
#| label: fig-DivRTT
#| out-height: 18cm
#| fig-pos: "H"
#| fig-cap: "Bayesian reconstructions of taxon longevities (A), diversification (B--C) and diversity (D) dynamics of South American Mammals from the early Eocene to the late Oligocene (genus level). The plot on the left (A) shows the longevity -- i.e., interval between origination and extinction times -- of each Eocene--Oligocene (373 out of 529 genera) South American mammal genus estimated with PyRate. J- and F-values respectively represent Jaccard's and Forbes' similarity indices computed between Eocene and Oligocene assemblages for each of the five groups. The right part of the panel shows the origination (blue) and extinction (red) rates through time (B), the net diversification rate (origination minus extinction) (C), and diversity (D) of all South American mammals throughout the study period. Full lines represent median estimates and ribbons their associated 95% higher posterior density. Light brown vertical bands indicate two key climate events, the Early Eocene Climate Optimum (EECO, ~53--51 Ma) and the Middle Eocene Climate Optimum (MECO, ~41.5--41 Ma). Stars on the (B) plot indicate significant rate shifts. Organisms' silhouettes are from PhyloPic and were added thanks to the R package rphylopic (98). Silhouettes were contributed by Julián Bayona, 2016 (CC BY-SA 3.0), Milena Cavalcanti, Patricia Pilatti & Diego Astúa (CC BY 4.0). Geological and South American Land Mammal Ages (SALMA) timescales were depicted thanks to the deeptime R package [@gearty2023]. Abbreviations for SALMA. Des: Deseadan, Ting: Tinguirirican, Must: Mustersan, Bar: Barrancan, Sap: “Sapoan Fauna”, R: Riochican, It: Itaboraian. The small band between Tinguirirican and Deseadan is the 'La Cantera fauna'."

include_graphics("../Chapter_1/figures/Main/Figure_1/FINAL_FIG_1.pdf")

```

### Ubiquitous taxonomic -- but not ecological -- turnover among the main mammalian groups across the study period

Our analyses of the overall SAM diversification dynamics failed at characterizing any mass extinction event at the time of the EOT, but recovered a late Eocene increase in mammal turnover (joint increase in origination and extinction rates), followed by an Oligocene waxing-and-waning diversity (@fig-DivRTT B and D). As a result, Eocene and Oligocene SAM appear taxonomically dissimilar. Indeed, the apparent proportion of shared genera between each temporal fauna, represented by the Jaccard's similarity index, is below 20% for four out of the five main groups we studied (@fig-DivRTT A and SI Appendix, Figure S38). Nevertheless, the corrected Forbes index, that was shown to perform more accurately when evaluating pairwise similarity between samples subject to data incompleteness [@alroy2015] reveals higher levels of faunal similarities, but still below 50% for three out of our five main studied groups (@fig-DivRTT A and SI Appendix, Figure S39). This does not call into question the occurrence of a taxonomic renewal, but suggests that it occurs at an attenuated magnitude than suggested by the Jaccard's index. The corrected genus longevity estimates corroborate results already established in the literature about mammal taxonomic replacement near the EOT, but also highlight previously undescribed patterns (@fig-DivRTT A and SI Appendix, Figure S18-21).

For notoungulates, the most taxon-rich order of our dataset (encompassing a third of our total occurrences; SI Appendix, Table S1), our inferences corroborate a long-recognized turnover at the order scale between the late Eocene and the early Oligocene [@croft2008]. This goes in line with a recent species-level analysis of southern South American notoungulate diversification dynamics, pointing toward an enhanced turnover in the early Oligocene, mostly among smaller-sized taxa [@solórzano2024]. According to our findings, this turnover is particularly striking at the family level, with the synchronous extinction and origination of three (Archaeopithecidae, Notostylopidae, and Oldfieldthomasiidae) and four families (Homalodontheriidae, Hegetotheriidae, Mesotheriidae, and Toxodontidae), respectively (SI Appendix, Figure S19-20). Most of these newly-arising families, in addition to two surviving ones (Archaeohyracidae and Interatheriidae), convergently evolved larger body sizes and higher teeth crown height (hypsodonty) than their basal counterparts, mostly during the Oligocene (though the first records of a hypsodont notoungulate lineage dates back to the middle Eocene, 20 million years before the first record of hypsodonty outside South America) [@solórzano2021; @stebbins1981]. As notoungulates likely represented a huge proportion of herbivorous lineages, these innovations likely implied changes in the architecture of the primary feeder mammalian guild between the Eocene and the Oligocene ecosystems at continent scale.

Considering the other native ungulate orders (i.e., Litopterna, Astrapotheria, and Pyrotheria), apart from the late Eocene extinction of Trigonostylopidae (Astrapotheria), evidence for higher-level taxonomic changes near the EOT is less obvious (SI Appendix, Figure S23). However, there seems to be a strong intra-family taxonomic renewal among most litopterns (i.e., Proterotheriidae, Macraucheniidae, and Adianthidae) and astrapotheriids (Astrapotheria) between the Eocene and Oligocene. These groups have been affected by a diversity decline from the late Eocene to the early Oligocene (SI Appendix, Figure S16D-F and S17D-F). Given that, just like notoungulates, these groups likely had herbivorous diets, the highlighted changes also surely affected the architecture of the ecosystems between the Eocene and the Oligocene.

In the case of metatherians (SI Appendix, Figure S18), our results are in line with previous findings about their taxonomic renewal near the EOT [@goin2016]. On one hand, entire superfamilies or sub-orders like Bonapartherioidea or Polydolopiformes (both subclades of the order Polydolopimorphia according to @goin2016) went extinct during the early Oligocene, meanwhile others, like Argyrolagoidea (also a subclade of Polydolopimorphia according to @goin2016), palaeothentoidea (Paucituberculata) and Hathliacynidae (Sparassodonta) originated and/or diversified. We may specify that the superfamily Palaeothentoidea appeared in the middle-late Eocene with the two stem genera Sasawatu and Perulestes, but the Palaeotenthidae and Abderitidae families diversified from the early Oligocene onward. On the other hand, intra-clade taxonomic replacement occurred among Borhyaenoidea (Sparassodonta) and Microbiotheria. As mentioned before, @goin2010 established that the most dramatic turnover among metatherian faunas occurred near the EOT (the 'Bisagra Patagónica'). Our corrected genus age reconstructions clearly bring qualitative support to such a macroevolutionary fate. However, when estimating metatherian diversification dynamics, no significant rate shift was found, and the overall number of genera through time remained stable (SI Appendix, Figure S17M-O). Quantitatively, our diversification analyses supported a constant turnover rate across the Eocene--Oligocene interval, which is likely a consequence of the low representation of metatherians in our dataset (219 occurrences, 54 genera).

As for cingulates, representing the majority of our xenarthran occurrences, we highlight a slowdown in Dasypodinae diversity, and the origin of Euphractinae in the middle-late Eocene (SI Appendix, Figure S21), in agreement with @ciancio2013. Just like for metatherians, our diversification reconstructions provided support for a constant-rate diversification throughout the period of interest, with a stable number of genera (SI Appendix, Figure S16I-L and 17I-L).

Regarding caviomorph rodents, the oldest representatives of the clade are found to originate in the end of the Eocene (@fig-DivRTT A and SI Appendix, Figure S22). Then, the group diversified at a constant rate until it reached a mid-Oligocene peak in extinction rate (SI Appendix, Figure S16G-I and S17G-I), and no main taxonomic replacement near the EOT is to be noticed.

To track the evolution of ecological composition between the Eocene and Oligocene SAM, we summarized trophic palaeocommunities by four dental ecomorphotype-derived categories: herbivorous, carnivorous, insectivorous and omnivorous. We used the origination and extinction ages estimated by PyRate to assess the number of genera through time within each ecological category and further computed the relative proportion of each trophic category at the geological sub-epoch level (SI Appendix, Figure S1 and Extended Materials and Methods). Our findings illustrate that the structure of trophic palaeocommunities remained stable throughout the entire time interval (@fig-Trophic). We report a slight increase in the relative proportion of herbivorous lineages in assemblages of decreasing age, at the expanse of omnivorous and insectivorous. In addition, ecological dissimilarity increases with the age difference between two assemblages, but remains low (below 25%), especially between the late Eocene and early Oligocene assemblages, when it equals 6.1% (@fig-Trophic).

Our results provide support for a taxonomic renewal between Eocene and Oligocene SAM faunas, with few changes in the architecture of ecosystems, thereby suggesting intra-guild renewal. Interestingly, the onset of this faunal turnover temporally matches the timing of 'the most fundamental biotic division in the Cenozoic' in the marine realm, that is, between the middle and the late Eocene [@prothero1994].

```{r, echo=FALSE}
#| label: fig-Trophic
#| out-width: 17cm
#| fig-cap: "Evolution of genus-level functional assemblages between the Eocene and the Oligocene. The plot on the left shows the median (full line) and 95% HPD (ribbon) of estimated genus richness among the four trophic classes we defined based on dental ecomorphotypes: carnivorous (red), herbivorous (green), insectivorous (purple) and omnivorous (blue). Pie charts indicate the relative abundance of each trophic group for each sub-epoch. The matrix on the right shows pairwise Bray-Curtis dissimilarities between sub-epoch-level trophic communities, with darker colour indicating higher dissimilarities. Geological and SALMA timescales were added thanks to the deeptime R package [@gearty2023]. Abbreviations for SALMA. Des: Deseadan, Ting: Tinguirirican, Must: Mustersan, Bar: Barrancan, Sap: “Sapoan Fauna”, Rio: Riochican. The small band between Tinguirirican and Deseadan is 'La Cantera fauna'."

include_graphics("../Chapter_1/figures/Main/Figure_2/Figure2.pdf")

```

### Latitudinal heterogeneity in diversification histories

Since the late break-up of the Gondwana supercontinent ca. 130 Mya, the South American landmass has offered a duality between tropical and non-tropical habitats [@scotese_atlas_2021]. This duality is particularly interesting as it allows for testing a long-standing question in macroevolution as to whether diversification processes occurred the same way inside and outside the tropics [@wallace1878]. Based on a time series of paleoclimate zone maps [@scotese2021; @quintero2023], we assigned each of our occurrences a tropical or extratropical affinity (*SI Appendix*, Figure S3), and analyzed genus-level diversification dynamics within each biome separately. Our findings revealed that extratropical genera exhibited diversity and diversification patterns extremely similar to those obtained from the complete dataset across the Eocene–Oligocene interval (Figure 3D-F and SI Appendix, Figure S8-11). Their diversity gradually declined from the middle Eocene to early Oligocene, with a late Eocene turnover increase (Figure 3D-F). This proximity between extratropical and continent-wide patterns was expected as extratropical taxa constitute the majority of genera present in our dataset (85.8%, SI Appendix, Table S10). Nevertheless, results are different when constraining our analyses to tropical genera (@fig-DivLat A--C). No significant change was found in the tropical origination rate, and one significant negative shift was found for the tropical extinction rate in the late Oligocene (but see later for an interpretation of this extinction peak) (@fig-DivLat A and S24). Their inferred diversity curve exhibits an opposite trend to the one obtained from extratropical genera. Tropical genera diversity increased exponentially during the Eocene and reached a plateau in the early Oligocene (@fig-DivLat C). In the late Oligocene, our reconstructions tend to suggest that tropical mammals experienced a diversity decline associated to the aforementioned increase in extinction rate. We propose this apparent decline to result from the scarcity of low-latitude fossil-bearing outcrops of Oligocene age, given that there are almost no tropical Deseadan (ca. 29.4--24.2 Ma) mammal remain unearthed to date (only two localities, CTA-32 and CTA-61 in the region of Contamana, Peru, see @antoine2017), therefore more reflecting a sampling bias rather than a true tropical SAM extinction in the late Oligocene. Consequently, after discarding this artefactual extinction peak, our results suggest that tropical lineages experienced no change in diversification rates, providing support for Wallace's theory of tropical stability hypothesis [@wallace1878]. Also, when considering attempts to derive early Oligocene ages to east-Amazonian occurrences originally assigned to an Eocene age [@campbell2021], the overall tropical diversity pattern remains unchanged. Indeed, the previously highlighted Eocene increase in diversity lasts a few more million years within the early Oligocene, until the onset of the 'decline' -- that we propose to be artificial (*SI Appendix*, Figure S33).

```{r, echo=FALSE}
#| label: fig-DivLat
#| fig-pos: "H"
#| fig-cap: Diversification (**A--B** and **D--E**) and diversity (**C** and **F**) dynamics of South American mammal genera between the Eocene and Oligocene arranged by paleotropical affinity. Trends are assessed for the tropical (**A--C**) and extratropical (**D--F**) biomes. Origination and extinction rates are respectively depicted in blue and red. Stars indicate significant rate shifts. The timing and significance of rate shifts was assessed by the rjMCMC model (Figure S36-37). Their difference, being the net diversification rate, is represented in black. Diversity through time is plotted in green. Light brown vertical bands indicate two key climate events, the Early Eocene Climate Optimum (EECO, ~53-51 Ma) and the Middle Eocene Climate Optimum (MECO, ~41.5-41 Ma). The question mark on the tropical diversity-through-time plot (**C**) indicates that the late-Oligocene tropical diversity decline that is observed is likely due to the quasi-absence of fossil material of late Oligocene age (Deseadan SALMA) so far unearthed. The maps on the right show the localities of our extratropical (top) and tropical (bottom) occurrences. Geological and SALMA timescales were added thanks to the deeptime R package [@gearty2023].

include_graphics("../Chapter_1/figures/Main/Figure_3/Figure3.pdf")
```

Our results suggest that highly distinct evolutionary regimes shaped tropical and extratropical mammal assemblages in South America. However, there is no clear evidence supporting commonly-assumed macroevolutionary hypotheses regarding how tropics shaped diversification processes (@fig-DivLat and *SI Appendix*, Figure S25-S29). Two macroevolutionary scenarios are usually invoked to explain the differences in diversity between tropical and extratropical biomes [@stebbins1974]. On one side, tropical lineages would have higher origination rates due to the higher diversity of habitats offered by the tropics ('cradle' hypothesis). On the other, tropical climate stability would make tropical lineages less prone to extinction ('museum' hypothesis, *but see* @vasconcelos2022). Here, tropical lineages appear to have a higher origination and a lower extinction rate than their extratropical relatives in the middle Eocene. This would typically be in line with [@rolland2014], who estimated temporal trends of speciation and extinction rates across latitudes using the mammal tree of life, and found that tropical mammals had a higher speciation and a lower extinction rate than their extratropical counterparts throughout the Cenozoic (*i*.*e*., 'cradle and museum' of diversity). More recently, an emphasis was put on higher-latitude tropical extinctions to explain the build-up of present-day tetrapod diversity [@meseguer2022; @quintero2023]. In alternative scenarios, varying dispersal rates can explain differences between tropical and extratropical diversity [@jablonski2006]. There is a consensus that tropical-to-extratropical transitions are prevalent, supporting the so-called 'out-of-the-tropics’ hypothesis [@jablonski2006], although extratropical-to-tropical transitions have been suggested [*e*.*g*., @meseguer2022; @smith2012; \textcolor{Blue}{Chapter 3}].

### Drivers behind the Eocene--Oligocene macroevolutionary dynamics of SAM

After exploring the dynamics of SAM diversity and diversification, we investigated their underlying drivers, addressing: what drove SAM diversity to decline in the 52--33.9 Ma interval? What caused their late Eocene increase in turnover? How do we explain such levels of taxonomic dissimilarity between Eocene and Oligocene SAM?

Our Multivariate Birth-Death (MBD) analyses indicate that the dissimilar Eocene and Oligocene faunas were ruled by different drivers (Figure 4). Overall, diversity-dependent effects largely impacted SAM diversification history throughout our study period, but these processes shaped both faunas differently. We recovered strong negative correlations of Eocene (post-EECO, 52--33.9 Ma) origination rate (correlation parameter $G_\lambda$ = -3.47, shrinkage weight \[$w_\lambda$\] = 0.91) and Oligocene extinction rate ($G_\mu$ = -2.36, $w_\mu$ = 0.83) with self-diversity (@fig-drivers, *SI Appendix*, Table S3--6). This suggests that Eocene mammal origination decreased when their diversity increased, which can be interpreted as a carrying capacity effect, that can occur at macroevolutionary scales [@storch2019]. The other diversity-dependent correlation indicates that the Oligocene mammal extinction rate decreased when their diversity increased. In the framework of the marked taxonomic turnover highlighted by our analyses, this effect might have arisen from facilitating interactions between recently-originated taxa (see forthcoming paragraph), but it could also be indirectly related to enhanced ecosystem productivity or niche partitioning events, which cannot be captured by the MBD analyses.

The MBD analyses also provided support for a positive correlation between global temperature and Eocene mammal origination rate ($G_\lambda$ = 0.24, $w_\lambda$ = 0.84). This explains the previously-described gradual decline in SAM diversity in the mid-late Eocene, qualitatively matching with the post-EECO overall decline in temperature (@fig-DivRTT D and @fig-drivers B). It is now clearly established that temperature change directly impacts biodiversity, at every timescale [@ipcc2023; @condamine2013; @malanoski2024], and temperature was already identified as a key driver influencing the diversification of an endemic European artiodactyl clade at the EOT [@weppe2021]. Based on isotope analyses of Patagonian ungulate enamel and on paleobotanical data, it has been proposed that the South American climate, at least in the Southern part of the continent, remained merely stable across the Eocene--Oligocene interval [@kohn2004; @kohn2015; @pound2017]. However, as illustrated by @antoine2021, we assume that the spatial scale (*i*.*e*., continental vs. global) does not affect the trends and directionality of temperature changes through time, but rather their amplitude. Hence, in the framework of our MBD analyses, the use of a global estimate of paleotemperature is still relevant, as we are more interested in trends than having actual paleotemperature values. Also, to date (*at the time the study was conducted*), no regional paleotemperature reconstruction for the South American continent has been produced with the same resolution as the global estimate used in this study [@westerhold2020].

We estimated a weak but significant positive correlation between mammal extinction rate and the Andean uplift both in the Eocene ($G_\lambda$ = 0.005, $w_\lambda$ = 0.68) and the Oligocene ($G_\mu$ = 0.02, $w_\mu$ = 0.83) (@fig-drivers, *SI Appendix*, Table S3-6). This Eocene correlation was maintained among high-latitude (extratropical) taxa when a more regionalised proxy of the pace of the Andean uplift was used (*SI Appendix*, Figure S41). Mountain-building processes are known to greatly influence species diversity [@antonelli2018], and previous studies illustrated the key importance of the Andes in shaping diversity patterns of a wide range of South American organisms, even in the early stages of the build-up of the chain [*e*.*g*., @antonelli2009; @boschman2022; @vallejos-garrido2023]. In a context of active margin, the Andean orogeny has been associated with volcanic activity, which, in conjunction to subsequent forest fires, resulted in a massive dust emission. This dust emission has been postulated to promote grazer-like herbivore origination and diversification, in particular at the EOT and thereafter [@kohn2015; @selkin2015]. Due to the high proportion of herbivores in our dataset (*SI Appendix*, Table S10 and @fig-Trophic), we could have expected Andean uplift to be either positively related to SAM origination, negatively related to SAM extinction, or both. Here, our results suggest that Andean uplift promoted extinction among SAM faunas, thus not providing support for any of the previously-stated hypotheses. Instead, they lead us to hypothesise that the building of the Andes, possibly by reshaping the landscapes locally, re-organising climate advections, triggering volcanism or any other factor, was an obstacle, rather than a facilitator, for SAM diversification across the Eocene--Oligocene interval.

Interestingly, when working with the complete dataset, none of our proxies for land opening, *i*.*e*., phytolith-derived floral composition -- proxy for grassland expansion, or habitat openness [@strömberg2013] -- and relative Leaf Area Index (rLAI) -- proxy for forest density [@dunn2015] -- were found to correlate with diversification rates. Similar conclusions were reached when restricting our analyses to notoungulates, the richest herbivorous clade of the dataset (*SI Appendix*, Figure S35). Nevertheless, fitting the MBD only to high-latitude (extratropical) taxa revealed a significant positive correlation of extinction rate with rLAI ($G_\mu$ = 2.56, $w_\mu$ = 0.96), and a negative correlation of origination rate with habitat openness ($G_\lambda$ = -0.06, $w_\lambda$ = 0.62) during the Oligocene (*SI Appendix*, Figure S40 and Table S11--S12). The interplay between grassland expansion and diversification of (herbivorous) mammal in South America has been debated in the literature [@pound2017; @stebbins1981; @strömberg2013]. The convergent evolution of high-crowned cheek teeth and larger body sizes in several notoungulate families near the EOT, suggesting the emergence of a grazing diet, has long been regarded as evidence for grassland expansion in response to climate cooling and aridification [@stebbins1981]. However, subsequent studies have shown that grasslands did not spread as much during the mid-Eocene to Oligocene of Patagonia, while hypsodont taxa evolved and diversified [@strömberg2013], and more generally that the composition of South American paleobiomes remained stable during this period [@pound2017]. Furthermore, trait-based correlation analyses failed to recover temporal relationship between the evolution of hypsodonty and the abundance of open habitats [@solórzano2021]. Here, our analyses provide contrasting evidence regarding the role of grassland spreading in SAM diversification. On a continental scale, no effect could be found, providing support against the common paradigm linking mammal diversification to grassland expansion. On the other hand, our result on a latitudinally-divided dataset revealed attenuated extinction and lower origination for Oligocene high-latitude grassland-dwelling taxa. It is currently difficult to determine whether this difference arises from an actual scale effect or heterogeneity in our data. Therefore, further research with increased sampling effort, especially at low latitudes, would be needed in order to address the question of the interplay between grassland expansion and SAM diversification.

When carrying out the MBD analyses without partitioning Eocene and Oligocene faunas, we still found strong negative correlations of origination ($G_\lambda$ = -2.16, $w_\lambda$ = 0.90) and extinction ($G_\mu$ = -2.18, $w_\mu$ = 0.81) rates with self-diversity, and a weak positive correlation of extinction rate with the Andean uplift ($G_\mu$ = 0.003, $w_\mu$ = 0.70) (*SI Appendix*, Figure S34, Table S3--6). However, we no longer retrieved any correlation of origination rate with temperature, highlighting the prominent and exclusive role of temperature in shaping Eocene mammal diversity.

```{r, echo=FALSE}
#| fig-label: fig-drivers
#| fig-cap: Bayesian estimation of correlation parameters on origination (blue) and extinction (red) rates with paleoenvironmental variables. (**A**) We analyzed separately the post-Early Eocene Climate Optimum (52--33.9 Ma, bottom row) and Oligocene (33.9--23.04 Ma, top row) mammal faunas, therefore not treating our dataset as a whole throughout our study period. The 2.5%, 50% and 97.5% quantiles are indicated in the distribution of each parameter by vertical lines, respectively from left to right. The interval between the leftmost and the rightmost vertical line of each parameter is therefore its 95% Highest Posterior Density (HPD). Significant correlations, *i*.*e*., having shrinkage weights higher than 0.5 and the correlation parameter not including 0 in their 95% HPD, are highlighted by stars and darker shades. Details of the environmental variables can be found in Figure S39 and Table S2 ('forest density’ refers to the relative Leaf Area Index, abbreviated rLAI). Values associated to coefficients and their corresponding shrinkage weights are available in *SI Appendix*, Table S3--S6. A graphical interpretation of the highlighted effects of environment covariates on South American mammal diversity is proposed in (**B**). Arrows with a sharp head indicate positive correlations, whereas flat heads indicate negative ones. Arrows with dotted lines indicate uncertain correlations, *i*.*e*., not retrieved from the analysis of the entire dataset but retrieved from the analysis of a latitudinal subsample of it (*SI Appendix*, Figure S41). Following color legend in (**A**), the arrow indicates the rate on which the correlation was found, with red being the extinction rate and blue the origination rate. Temperature during the Eocene (displayed in pink) was taken from @westerhold2020. The mountain drawing refers to the Andean uplift, and the grass drawing refers to the proportion of open habitats (amalgamated with forest density).

include_graphics("../Chapter_1/figures/Main/Figure_4/FULL_FIG_4.pdf")

```

## Materials and Methods

Additional methodological details are provided in the *SI Appendix* to this chapter.

### Fossil data compilation and analytical framework

We downloaded and cleaned all genus-level South American terrestrial mammal occurrences from The Paleobiology Database <https://paleobiodb.org/> spanning the late Paleocene to the early Miocene (60--16 Ma) (3,384 occurrences; *SI Appendix, Extended Materials and Methods* and Tab. S1). Cleaning consisted of a manual revision of the taxonomic name (*e*.*g*., synonymies, missing open nomenclature elements, typos) and time range of each occurrence (*SI Appendix*, Figure S4 and *Manual Cleaning Details*). SAM fossil data are prone to a large spatial heterogeneity, with most fossil localities located close to or high in the Andes -- in their current conformation (*SI Appendix*, Figure S3). Using palaeocoordinate reconstructions [@scotese2018; @jones2023] and paleobiome maps [@scotese2021], we assigned each occurrence a tropical or extratropical affinity, and collapsed their associated dental ecomorphotype-derived trophic classes into four categories: carnivorous, herbivorous, omnivorous and insectivorous (*SI Appendix*, Tab. S10).

We analyzed our database using PyRate 3 [@silvestro_improved_2019] and approximated the posterior distribution of the origination and extinction times ($T_s$ and $T_e$, respectively) of each genus and the rate of the preservation process using the rjMCMC algorithm. We then used these $T_s$ and $T_e$ to estimate the underlying diversification process (origination and extinction rates) with the Birth-Death model with Constrained Shifts (BDCS) [@silvestro_revisiting_2015] (*SI Appendix, Extended Materials and Methods*).

We ran the rjMCMC model for 50 million generations and sampled every 50,000, with a time-dependent preservation process allowing for rate shift between geological stages (*SI Appendix*, Figure S5-6 and Extended Materials and Methods). We allowed for a gamma-distributed preservation heterogeneity across lineages. We ran the BDCS model for 10 million generations, sampled every 10,000, and allowed diversification (*i*.*e*., origination and extinction) rates to vary between each geological sub-Epochs (*SI Appendix*, Figure S1). Diversification analyses were replicated 20 times after randomly drawing fossil ages between their upper and lower age boundaries, and posterior estimates across all replicates that achieved convergence ($n = 11$, *SI Appendix*, Figure S42) were combined. To test the robustness of our macroevolutionary inference, we conducted sensitivity analyses by either removing singletons or subsampling our fossil database to unique locality-lineage combinations (*SI Appendix, Extended Materials and Methods*). We assessed the convergence of each parameter with the pymc 5.9.2 python library [@wiecki2023]. We considered parameters achieved convergence if their Effective Sample Size (ESS) was above 200.

### Age-dependent Extinction (ADE) model

We assessed the effect of taxon age on its extinction probability by fitting the ADE model to our genus-level occurrence data [@hagen2018]. We ran the model for 100 million generations and sampled every 50,000, with preservation rates varying between geological stages and across lineages according to a Gamma model. We focused on the shape ($\Phi$) of the Weibull distribution, knowing that $\Phi<1$ indicates a decreasing extinction probability with increasing taxon age and $\Phi>1$ indicates an increasing extinction probability with increasing taxon age (no effect if $\Phi=1$). $\Phi$ was considered as significantly different from one if one was excluded from a significant proportion of its posterior distribution. We performed 20 independent ADE analyses of randomised datasets before and after the increase in turnover to assess the age selectivity of this taxonomic renewal (*SI Appendix*, Figure S8-11, S43 and *Extended Materials and Methods*). We monitored parameter convergence with the pymc 5.9.2 python library, and parameters were considered convergent if their ESS was above 200.

### Selection of environmental variables

To investigate possible environmental correlates for the changes in SAM diversification dynamics over the Eocene−Oligocene interval, we selected two global (*i*.*e*., atmospheric temperature and sea level) and four continental-scale (*i*.*e*., Andean uplift, Neotropical plant diversity, forest density and habitat openness) variables (*SI Appendix*, Figure S40 and Tab. S2). Temperature is known for influencing macroevolutionary processes across several groups, including mammals [@condamine2019; @barnosky2001], and sea level can impact the connectivity of habitats. Even in its early phase, the Andean uplift was evidenced as a major diversification driver for many South American groups [@boschman2022]. Diversity of SAM, in particular herbivorous taxa, can be controlled by plant diversity. Finally, relative Leaf Area Index (proxy for forest density, @pound2017) and patterns of relative abundance of open habitats [@solórzano2021; @strömberg2013; @palazzesi2012] are both indicators of the relative abundance of grasslands, which has long been regarded as shaping herbivore diversity [@stebbins1981]. As highlighted by @weppe2023, it is important to consider regional-scale variables because environmental changes across the Eocene--Oligocene interval were shown to be regionally heterogeneous [@pound2017; @toumoulin2022].

### Multivariate Birth-Death (MBD) model

We carried out environment-dependent diversification analyses involving the aforementioned variables by running the MBD model included in PyRate [@lehtonen2017]. This model carries out two independent correlations involving a set of environmental covariates and diversification rates ($\lambda$ and $\mu$) newly-estimated from empirical $T_s$ and $T_e$. Each variable is assigned to two correlation coefficients, one for the correlation with $\lambda$ ($G_{i, \lambda}$, where $i$ is the index of the variable in the set of covariates) and the other for the correlation with $\mu$ ($G_{i, \mu}$). An MCMC algorithm jointly estimates baseline origination and extinction rates and all the Gλ and Gµ, while controlling for over-parameterization. Each $G_\lambda$ and $G_\mu$ is assigned a shrinkage weight ($w(G_\lambda)$ or $w(G_\mu)$), ranging between 0 and 1, testing the significance of the correlation. A correlation of coefficient $G_{i, \lambda}$ is considered significant if $w(G_{\lambda, i}) > 0.5$ and if a sufficiently high proportion of its posterior density does not overlap with 0. Each set of MBD analyses was run across 10 million iterations, sampling every 10,000. We carried out exponential correlations. As we expected Eocene and Oligocene faunas to be influenced by different drivers, we also performed independent MBD analyses for the Eocene (post-EECO, *i*.*e*., 52--33.9 Ma) and for the Oligocene (33.9--23.03 Ma) taxa (*SI Appendix*, Figure S44). Parameter convergence (*i*.*e*., $ESS > 200$) was assessed using pymc. 

## Data, Materials, and Software Availability 

All the occurrence data used in this study are available on a Figshare repository (<https://doi.org/10.6084/m9.figshare.26495989>) and all the data analysis and visualization codes have been deposited in GitHub (<https://github.com/Buffan3369/Neotropical_mammals.git> and <https://github.com/Alexis-Marion/CorsaiR.git>).

### Acknowledgments

We are grateful to all the people who collected the data that we used in the field. We are also extremely grateful toward Evangelos Vlachos, Jelle S. Zilstra, John Alroy, Philip D. Mannion, Carlos Jaramillo, Darin Croft, Mark D. Uhen, Richard J. Butler, Matthew T. Carrano, Patricia Holroyd, Ascanio D. Rincón, Graeme T. Lloyd, Luciano L. Rasia and Jason Head for entering these data in The Paleobiology Database. Moreover, we thank Chris Scotese (Northwestern Univ., Evanston, USA), Ignacio Quintero (IBENS, France) and Miranta Kouvari for their data sharing; Gilles Escarguel (LEHNA, Lyon, France), Myriam Boivin (CONICET, San Salvador de Jujuy, Argentine) and Ana Rodrigues (CEFE, Montpellier, France) for their useful feedbacks on the present study. Also, we thank Alexis F.P. Marion (ISEM, France) for coding collaborations and thoughtful remarks; and all members of the Phylogénie et Évolution Moléculaire and Paléontologie teams at ISEM for their useful inputs and feedbacks on the present work. Additionally, we thank Nils C. Stenseth for handling our work as a NAS editor, Camilo Montes and two anonymous reviewers for their impactful reviews on the present study. This research was supported by the French Agence Nationale de la Recherche (ANR) in the framework of the LabEx CEBA (ANR-10-LABX-25-01; strategic project EMERGENCE). L.B. was funded by a grant offered by the École Normale Supérieure de Lyon. N.S. was funded by the Brazil Conselho Nacional de Desenvolvimento Científico e Tecnológico (CNPq) grant nº 443571/2023-6. This is Paleobiology Database publication no. 524 and ISEM publication no. 2025-062 SUD. “Quel moyen de paiement utilisaient les mammifères à la transition Éocène−Oligocène ? Les Grandes Coupures !”
