# Chapter 1. The fate of South America's endemic mammalian fauna in response to the most dramatic Cenozoic climate disruption {.unnumbered header-title="Chapter 1"}

## Foreword

```{r, echo=FALSE}
library(knitr)
```

Understanding what drove the emergence and demise of lineages during biodiversity crises may provide key insights at the time of the sixth mass extinction. Here, using a vetted fossil database, we explored how South American mammals (SAM) responded to the most dramatic Cenozoic climate cooling episode, the Eocene−Oligocene transition (ca. 34 Ma). We found that SAM experienced a gradual and long-standing diversity decline strongly related to late Eocene progressive climate cooling, followed by a major taxonomic turnover in the Oligocene, driven by diversity-dependent factors, Andean uplift and perhaps grassland expansion. SAM were therefore more subject to long-term diversity changes rather than abrupt ones, contrary to what would be expected in a mass extinction scenario.

The content of this chapter was published in PNAS [@buffan2025].

## Abstract

Around 34 million years ago (Ma), the Eocene−Oligocene transition (EOT) marked the most dramatic global climatic cooling of the Cenozoic. On a planetary scale, palaeontological evidence suggests that this transition was associated with major faunal turnovers, sometimes even regarded as a mass extinction crisis. In South America, there is no consensus on the response of the endemic mammals to this transition. Here, using a vetted fossil dataset and cutting-edge Bayesian methods, we analyzed the dynamics of South American mammal (SAM) diversification and their possible drivers across latitude (tropical vs. extratropical), taxonomic groups, and trophic guilds throughout the Eocene−Oligocene (ca. 56−23 Ma). Our results did not evidence any mass extinction among SAM at the EOT. Instead, they experienced a gradual and long-term diversity decline from the middle Eocene to the early Oligocene, followed by a sudden waxing-and-waning diversity associated with a large taxonomic -- but not ecological -- turnover. Tropical and extratropical lineages have had very distinct macroevolutionary histories. No effective change in the pace at which tropical lineages diversify was found, thus favoring the tropical stability hypothesis proposed by Wallace. Diversity-dependent effects, temperature and Andean uplift were recovered as probable drivers of SAM diversification across the period. Contrasting evidence casts doubt on the common hypothesis primarily linking Oligocene faunal changes to grassland expansion. Our findings illustrate the uniqueness of the deep-time interplay between endemic SAM and their physical environment in a context of climatic shift, highlighting the need to consider regional idiosyncrasies for understanding the coevolution of life and climate.

## Introduction

Today, South America is home to a unique mammalian fauna, hosting one of the most species-rich regions of the world, Amazonia [@willig2003]. Although the extant mammalian taxonomic diversity in the continent is astonishing, it is a poor reflection of what can be drawn from the fossil record. In fact, throughout most of the Cenozoic (i.e., the last 66 million years), South America was isolated from other landmasses [@scotese_atlas_2021]. In this insular context, terrestrial mammal lineages that inhabited this huge landmass experienced unique radiations [@goin2016; @boivin2019; @croft2020; @beck2023; @tejada2024]. According to @simpson1980 , South American mammal (SAM) communities through the Cenozoic are well summarized by three successive ‘faunal strata’. The first one, stretching from the early palaeocene (66 million years ago, Ma) to the late Eocene (ca. 37 Ma), is characterized by the early establishment and radiation of endemic lineages, mostly xenarthrans, metatherians and native ungulates. Then, from the early Oligocene (ca. 33 Ma) until the Late Miocene-Early Pliocene (ca. 5 Ma), key morphological innovations appeared among native lineages, leading to a so-called ‘faunal modernization’ [@croft2008]. In the meantime, rodents and primates, Afro-Asian lineages that most likely reached South America in the middle-late Eocene [@antoine2012; @bond2015; @marivaux2023], diversified locally [@boivin2019; @beck2023]. In Plio-Pleistocene times, this endemic fauna was massively disrupted. Meanwhile, the closure of the Panama isthmus [@montes2015] had paved the way for major faunal exchanges between North and South America, described as the ‘Great American Biotic Interchange’ [@marshall1988], and immigrant taxa from North America (*e*.*g*., carnivores, perissodactyls, artiodactyls, or proboscideans) appeared in the South American fossil record. In the last millions of years, many endemic South American lineages declined, or even went totally extinct (e.g., litopterns, notoungulates, sparassodonts, glyptodonts, ground sloths).

Close to the limit between Simpson’s first and second faunal strata, the Eocene−Oligocene transition (EOT, *ca*. 33.9 Ma) marked the most dramatic Cenozoic climatic change episode worldwide [@coxall2007]. At that time, simultaneously, global climate massively cooled down [@westerhold2020], ocean level dropped [@miller2020], and the whole Antarctic continent turned into a giant ice sheet [@liu2009]. The prevailing mechanism proposed to explain these major environmental changes is the onset of the Antarctic circumpolar current [@toumoulin2020]. Concomitantly, a major biotic turnover among mammals, known as the '*Grande Coupure*', occurred in Europe [@stehlin1909]. This event, punctuated by the arrival of immigrant taxa from Asia, saw the demise of many endemic European lineages [@legendre1991; @escarguel2008; @weppe2023]. Although European-oriented case studies dominate the literature about the topic, similar disruptions among mammalian faunas occurred at the same time in other regions of the world, such as Asia [@meng1998; @kraatz2010; @sun_synchronous_2014], North America [@figueirido2012; @stucky2014], Africa [@devries2021] or to some extent Madagascar [@godfrey2020]. Recently, @hoyalcuthill2020 even came to rank this geological boundary as the second-most important extinction event throughout the history of life.

In central and southern South America, substantial changes have been documented in metatherians [@abello2020; @goin2010] and native ungulates [@croft2020; @solórzano2024] near the EOT. As an echo to the '*Grande Coupure*', @goin2010 even introduced the term '*Bisagra Patagónica*' (Patagonian Hinge) to characterize the changes among metatherian communities in Patagonia possibly related to the EOT. General morphological trends have been highlighted in 'modernised' notoungulate faunas, with a net increase in body mass and cheek teeth’s crown height [@solórzano2021]. The latter changes, by analogy with the evolutionary fate of extant groups, suggested the emergence of the grazing feeding mode, and was postulated to result from the expansion of grasslands, promoted by post-EOT climate cooling and aridification [@solórzano2024; @stebbins1981; @reguero2010]. In northern South America, changes in methaterian and rodent communities around the EOT have also been suggested despite the paucity of known Palaeogene low-latitude fossil localities to date [@antoine2021]. Additionally, geological, palaeontological and phylogenetic evidence suggests that the Eocene−Oligocene was a key interval for the dispersal of SAM toward Caribbean Islands, further playing a central role in the architecture of these unique insular ecosystems [@tejada2024; @philippon2020; @cornée2021; @marivaux_early_2020].

Despite general agreement that a faunal remodelling occurred in SAM communities near the EOT, the evolutionary tempo and mode remain unknown. At macroevolutionary scales, as we would expect in a mass extinction scenario (*SI Appendix*, *Supplementary text*), it is unclear whether this faunal renewal was characterized by extinction of Eocene lineages followed by diversification of the surviving ones [@guinot2023], or whether lineages gradually transitioned until they became 'modernised'. A crucial question underlying these conflicting macroevolutionary hypotheses is whether or not extinction occurred in the tropics, and at what level. Indeed, it has long been postulated that the climatic stability of tropics would make tropical lineages less vulnerable to extinction than extratropical lineages [@wallace1878]. Also, the ecological selectivity and the likely drivers of this apparent faunal transition have remained underexplored. For instance, there is currently no consensus regarding the role of temperature [@kohn2004; @kohn2015] or grassland expansion [@strömberg2013] in shaping the macroevolutionary fate of SAM. Finally, there is a lack of integrative frameworks at the continental scale, with most case studies at the local scale, mostly in southern South America. Nevertheless, the Palaeogene mammal fossil record in South America is spatially and temporally uneven. Spatial gaps cover large areas, such as most of the Amazon basin, and important time intervals have remained unsampled to date (*SI Appendix*, *Fig. S1* and *S3*). Accordingly, it has been historically challenging to address questions regarding the history of SAM at the continental scale.

Here, we aim to fill these gaps by examining the available fossil record of SAM. We assembled a genus-level fossil dataset consisting of 3,378 occurrences (distributed in 1,115 species, 536 genera, 96 families, and 18 orders) spanning from the late palaeocene to the Early Miocene. Relying on a fossil-based Bayesian birth-death framework accounting for variable preservation rates through time and across taxa [@silvestro_bayesian_2014], we estimated the tempo and modes of the mammalian diversification dynamics in South America throughout the Eocene--Oligocene interval (*ca*. 56--23 Ma). We then disentangled the underlying taxonomic, functional, and geographic (tropical vs. extratropical) bases of the inferred continent-scale patterns of diversity and diversification among SAM faunas. We finally quantified the most likely biotic and abiotic drivers explaining their diversification dynamics.

## Results and discussion

### No mass extinction: South American mammal diversity gradually declined well before the EOT

We analyzed the fossil database (Tab. S1) both at the genus and species levels using the process-based Bayesian framework of diversification accounting for varying preservation implemented in PyRate [@silvestro_bayesian_2014]. Our results highlight that the overall mammal diversity gradually declined from the end of the early Eocene -- shortly after the Early Eocene Climate Optimum (EECO, 45) -- to the end of the early Oligocene (decline by respectively 34% and 52% in the number of genera and species between 48 and 32 Ma), and then further increased in the late Oligocene (@fig-DivRTT D and SI Appendix, Fig. S8-11). No massive loss of lineages was evidenced at the EOT. This diversity trajectory results from a significant decrease in origination rate during the early Eocene (@fig-DivRTT B and SI Appendix, Fig. S8-11). Additionally, we recover a joint significant increase in the pace of origination and extinction (i.e., increased turnover) during the late Eocene, with the extinction rate always being higher than the origination rate (slightly negative net diversification rate; @fig-DivRTT B--C and SI Appendix, S8-11). Similar diversification patterns are recovered when analyzing the dataset at the genus or species level (SI Appendix, Fig. S8-11), after removing singletons (SI Appendix, Fig. S12-13) or disabling several occurrences of the same lineage that co-occur temporally and geographically (SI Appendix, Fig. S14-15). Moreover, general trends in diversity and diversification dynamics are conserved when working with occurrences assigned to South American Land Mammal Ages (SALMAs; SI Appendix, Fig. S9 and S11) instead of sub-epochs (see Methods; SI Appendix, Fig. S1-2 and Extended Material and Methods), or when considering attempts to derive early Oligocene ages to western Amazonian occurrences originally assigned to an Eocene age (SI Appendix, Fig. S32).

Recently, an analysis of the global Phanerozoic fossil record classified the EOT as the second most intense mass extinction episode [@hoyalcuthill2020] (see SI Appendix, Supplementary text for a definition of a mass extinction). Although this ranking was unexpectedly high, drastic extinctions at the EOT are supported by analyses of the European [@stehlin1909] and Asian [@meng1998] mammal fossil record. Applying a methodological framework similar to ours, @weppe2023 recovered a strong extinction signal among endemic Western European artiodactyls at the EOT, with the loss of 62% of genera and 77% of species in only 1 My. Other studies working on mass extinctions illustrated similarly high levels of lineage loss for various study models, for instance in insects at the Permian/Triassic boundary (82% genus loss) [@jouault_multiple_2022] or in elasmobranchs – sharks, skates and rays – at the Cretaceous/palaeogene boundary (44.94% genus and 62.6% species loss) [@guinot2023]. Here, given the relatively small number of lineages lost at the Eocene−Oligocene boundary − 15 genera (19%), 19 species (22%) −, our results are not consistent with a mass extinction event among SAM at the EOT. Instead, we highlight a long-term decline in SAM diversity, from the middle Eocene onward, contrasting with the expected short temporality of a mass extinction crisis. To our knowledge, our work is the first to provide quantitative support for such a pattern in SAM. These differences between South America and other continents could come from the tropical nature of and the ecological continuum within a significant part of the landmass. Stemming from the cornerstone hypothesis of Wallace linking the striking similarities between tropical biomes from all-over the world to the uniformity of tropical climate across space and (shallow) time [@wallace1878], tropical climate stability would act as a buffer for tropical lineages against extinction. Also, the isolated condition of the landmass has been proposed as a probable cause explaining the singularity of SAM near-EOT macroevolutionary fate, as it greatly limited the possibilities for immigration events. Yet, immigrations were hypothesized to have played a key role in EOT-related extinctions in other parts of the world [@goin2016] [but see @escarguel2008].

```{r, echo=FALSE}
#| label: fig-DivRTT
#| out-height: 18cm
#| fig-pos: "H"
#| fig-cap: "Bayesian reconstructions of taxon longevities (A), diversification (B--C) and diversity (D) dynamics of South American Mammals from the early Eocene to the late Oligocene (genus level). The plot on the left (A) shows the longevity -- i.e., interval between origination and extinction times -- of each Eocene--Oligocene (373 out of 529 genera) South American mammal genus estimated with PyRate. J- and F-values respectively represent Jaccard’s and Forbes’ similarity indices computed between Eocene and Oligocene assemblages for each of the five groups. The right part of the panel shows the origination (blue) and extinction (red) rates through time (B), the net diversification rate (origination minus extinction) (C), and diversity (D) of all South American mammals throughout the study period. Full lines represent median estimates and ribbons their associated 95% higher posterior density. Light brown vertical bands indicate two key climate events, the Early Eocene Climate Optimum (EECO, ~53--51 Ma) and the Middle Eocene Climate Optimum (MECO, ~41.5--41 Ma). Stars on the (B) plot indicate significant rate shifts. Organisms’ silhouettes are from PhyloPic and were added thanks to the R package rphylopic (98). Silhouettes were contributed by Julián Bayona, 2016 (CC BY-SA 3.0), Milena Cavalcanti, Patricia Pilatti & Diego Astúa (CC BY 4.0). Geological and South American Land Mammal Ages (SALMA) timescales were depicted thanks to the deeptime R package [@gearty2023]. Abbreviations for SALMA. Des: Deseadan, Ting: Tinguirirican, Must: Mustersan, Bar: Barrancan, Sap: “Sapoan Fauna”, R: Riochican, It: Itaboraian. The small band between Tinguirirican and Deseadan is the 'La Cantera fauna'."

include_graphics("../Chapter_1/figures/Main/Figure_1/FINAL_FIG_1.pdf")

```

### Ubiquitous taxonomic -- but not ecological -- turnover among the main mammalian groups across the study period

Our analyses of the overall SAM diversification dynamics failed at characterizing any mass extinction event at the time of the EOT, but recovered a late Eocene increase in mammal turnover (joint increase in origination and extinction rates), followed by an Oligocene waxing-and-waning diversity (@fig-DivRTT B and D). As a result, Eocene and Oligocene SAM appear taxonomically dissimilar. Indeed, the apparent proportion of shared genera between each temporal fauna, represented by the Jaccard’s similarity index, is below 20% for four out of the five main groups we studied (@fig-DivRTT A and SI Appendix, Fig. S38). Nevertheless, the corrected Forbes index, that was shown to perform more accurately when evaluating pairwise similarity between samples subject to data incompleteness [@alroy2015] reveals higher levels of faunal similarities, but still below 50% for three out of our five main studied groups (@fig-DivRTT A and SI Appendix, Fig. S39). This does not call into question the occurrence of a taxonomic renewal, but suggests that it occurs at an attenuated magnitude than suggested by the Jaccard’s index. The corrected genus longevity estimates corroborate results already established in the literature about mammal taxonomic replacement near the EOT, but also highlight previously undescribed patterns (@fig-DivRTT A and SI Appendix, Fig. S18-21).

For notoungulates, the most taxon-rich order of our dataset (encompassing a third of our total occurrences; SI Appendix, Tab. S1), our inferences corroborate a long-recognized turnover at the order scale between the late Eocene and the early Oligocene [@croft2008]. This goes in line with a recent species-level analysis of southern South American notoungulate diversification dynamics, pointing toward an enhanced turnover in the early Oligocene, mostly among smaller-sized taxa [@solórzano2024]. According to our findings, this turnover is particularly striking at the family level, with the synchronous extinction and origination of three (Archaeopithecidae, Notostylopidae, and Oldfieldthomasiidae) and four families (Homalodontheriidae, Hegetotheriidae, Mesotheriidae, and Toxodontidae), respectively (SI Appendix, Fig. S19-20). Most of these newly-arising families, in addition to two surviving ones (Archaeohyracidae and Interatheriidae), convergently evolved larger body sizes and higher teeth crown height (hypsodonty) than their basal counterparts, mostly during the Oligocene (though the first records of a hypsodont notoungulate lineage dates back to the middle Eocene, 20 million years before the first record of hypsodonty outside South America) [@solórzano2021; @stebbins1981]. As notoungulates likely represented a huge proportion of herbivorous lineages, these innovations likely implied changes in the architecture of the primary feeder mammalian guild between the Eocene and the Oligocene ecosystems at continent scale.

Considering the other native ungulate orders (i.e., Litopterna, Astrapotheria, and Pyrotheria), apart from the late Eocene extinction of Trigonostylopidae (Astrapotheria), evidence for higher-level taxonomic changes near the EOT is less obvious (SI Appendix, Fig. S23). However, there seems to be a strong intra-family taxonomic renewal among most litopterns (i.e., Proterotheriidae, Macraucheniidae, and Adianthidae) and astrapotheriids (Astrapotheria) between the Eocene and Oligocene. These groups have been affected by a diversity decline from the late Eocene to the early Oligocene (SI Appendix, Fig. S16D-F and S17D-F). Given that, just like notoungulates, these groups likely had herbivorous diets, the highlighted changes also surely affected the architecture of the ecosystems between the Eocene and the Oligocene.

In the case of metatherians (SI Appendix, Fig. S18), our results are in line with previous findings about their taxonomic renewal near the EOT [@goin2016]. On one hand, entire superfamilies or sub-orders like Bonapartherioidea or Polydolopiformes (both subclades of the order Polydolopimorphia according to @goin2016) went extinct during the early Oligocene, meanwhile others, like Argyrolagoidea (also a subclade of Polydolopimorphia according to @goin2016), palaeothentoidea (Paucituberculata) and Hathliacynidae (Sparassodonta) originated and/or diversified. We may specify that the superfamily Palaeothentoidea appeared in the middle-late Eocene with the two stem genera Sasawatu and Perulestes, but the Palaeotenthidae and Abderitidae families diversified from the early Oligocene onward. On the other hand, intra-clade taxonomic replacement occurred among Borhyaenoidea (Sparassodonta) and Microbiotheria. As mentioned before, @goin2010 established that the most dramatic turnover among metatherian faunas occurred near the EOT (the ‘Bisagra Patagónica’). Our corrected genus age reconstructions clearly bring qualitative support to such a macroevolutionary fate. However, when estimating metatherian diversification dynamics, no significant rate shift was found, and the overall number of genera through time remained stable (SI Appendix, Fig. S17M-O). Quantitatively, our diversification analyses supported a constant turnover rate across the Eocene--Oligocene interval, which is likely a consequence of the low representation of metatherians in our dataset (219 occurrences, 54 genera).

As for cingulates, representing the majority of our xenarthran occurrences, we highlight a slowdown in Dasypodinae diversity, and the origin of Euphractinae in the middle-late Eocene (SI Appendix, Fig. S21), in agreement with @ciancio2013. Just like for metatherians, our diversification reconstructions provided support for a constant-rate diversification throughout the period of interest, with a stable number of genera (SI Appendix, Fig. S16I-L and 17I-L).

Regarding caviomorph rodents, the oldest representatives of the clade are found to originate in the end of the Eocene (@fig-DivRTT A and SI Appendix, Fig. S22). Then, the group diversified at a constant rate until it reached a mid-Oligocene peak in extinction rate (SI Appendix, Fig. S16G-I and S17G-I), and no main taxonomic replacement near the EOT is to be noticed.

To track the evolution of ecological composition between the Eocene and Oligocene SAM, we summarized trophic palaeocommunities by four dental ecomorphotype-derived categories: herbivorous, carnivorous, insectivorous and omnivorous. We used the origination and extinction ages estimated by PyRate to assess the number of genera through time within each ecological category and further computed the relative proportion of each trophic category at the geological sub-epoch level (SI Appendix, Fig. S1 and Extended Materials and Methods). Our findings illustrate that the structure of trophic palaeocommunities remained stable throughout the entire time interval (@fig-Trophic). We report a slight increase in the relative proportion of herbivorous lineages in assemblages of decreasing age, at the expanse of omnivorous and insectivorous. In addition, ecological dissimilarity increases with the age difference between two assemblages, but remains low (below 25%), especially between the late Eocene and early Oligocene assemblages, when it equals 6.1% (@fig-Trophic).

Our results provide support for a taxonomic renewal between Eocene and Oligocene SAM faunas, with few changes in the architecture of ecosystems, thereby suggesting intra-guild renewal. Interestingly, the onset of this faunal turnover temporally matches the timing of ‘the most fundamental biotic division in the Cenozoic’ in the marine realm, that is, between the middle and the late Eocene [@prothero1994].

```{r, echo=FALSE}
#| label: fig-Trophic
#| out-width: 17cm
#| fig-label: "Evolution of genus-level functional assemblages between the Eocene and the Oligocene. The plot on the left shows the median (full line) and 95% HPD (ribbon) of estimated genus richness among the four trophic classes we defined based on dental ecomorphotypes: carnivorous (red), herbivorous (green), insectivorous (purple) and omnivorous (blue). Pie charts indicate the relative abundance of each trophic group for each sub-epoch. The matrix on the right shows pairwise Bray-Curtis dissimilarities between sub-epoch-level trophic communities, with darker colour indicating higher dissimilarities. Geological and SALMA timescales were added thanks to the deeptime R package [@gearty2023]. Abbreviations for SALMA. Des: Deseadan, Ting: Tinguirirican, Must: Mustersan, Bar: Barrancan, Sap: “Sapoan Fauna”, Rio: Riochican. The small band between Tinguirirican and Deseadan is 'La Cantera fauna'."

include_graphics("../Chapter_1/figures/Main/Figure_2/Figure2.pdf")

```

